Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-16 paper/paper.tex
start discussion of less plausible explanations

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence indicating that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref[fig:tree_RaTG13} of simply placing the root at the three possible locations that are closest to the bat outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} or my more simplistic approach in Figure~\ref[fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

@@

>>>>

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which is reported to have been collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.

Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the three possible locations closest to the bat outgroup.
However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.

Before moving onto the next section, I will also briefly address several less plausible explanations for the discordance that have gained traction in popular discussion of SARS-CoV-2's origins.
The first explanation suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
This explanation can be discounted, because although there are some unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the fundamental conundrum about the progenitor described above holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before the emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.

@@

\subsection{Sequences recovered from the deleted project help reconcile inferences about SARS-CoV-2's progenitor}
